% ------------------------------------------------------------------
\documentclass[12 pt]{article} % A4 paper set by geometry package below
\pagenumbering{arabic}
\setlength{\parindent}{10 mm}
\setlength{\parskip}{12 pt}

% Nimbus Sans font should be reasonably legible
\usepackage{helvet}
\renewcommand{\familydefault}{\sfdefault}
\usepackage[T1]{fontenc}  % Without this \textsterling produces $

% Section header spacing
\usepackage{titlesec}
\titlespacing\section{0pt}{12pt plus 4pt minus 2pt}{0pt plus 2pt minus 2pt}
\titlespacing\subsection{0pt}{12pt plus 4pt minus 2pt}{0pt plus 2pt minus 2pt}
\titlespacing\subsubsection{0pt}{12pt plus 4pt minus 2pt}{0pt plus 2pt minus 2pt}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{verbatim}    % For comment
\usepackage[shortlabels]{enumitem}
\usepackage[paper=a4paper, marginparwidth=0 cm, marginparsep=0 cm, margin=2.5 cm, includemp]{geometry}
\usepackage[pdftex, pdfstartview={FitH}, pdfnewwindow=true, colorlinks=true, citecolor=blue, filecolor=blue, linkcolor=blue, urlcolor=blue, pdfpagemode=UseNone]{hyperref}

% Put module code and last-modified date in footer
\usepackage{fancyhdr}
\pagestyle{fancy}
\fancyhf{}
\renewcommand{\headrulewidth}{0pt}
\cfoot{{\small \thisunit}\hfill \thepage\hfill {\small \moddate}}

% Hopefully address Canvas complaints about pdf tagging
%\usepackage[tagged]{accessibility}
\hypersetup {
  pdfauthor={David Schaich},
  pdftitle={StatMech Extra Practice},
}
% ------------------------------------------------------------------



% ------------------------------------------------------------------
% Shortcuts
\newcommand{\De}{\ensuremath{\Delta} }
\newcommand{\si}{\ensuremath{\sigma} }
\renewcommand{\d}[1]{\ensuremath{\mathop{d#1}} }
% ------------------------------------------------------------------



% ------------------------------------------------------------------
\begin{document}
\newcommand{\thisunit}{MATH327 Extra (Inventory)}
\newcommand{\moddate}{Last modified 27 Feb.~2025}
\begin{center}
  {\Large \textbf{MATH327: StatMech and Thermo, Spring 2025}} \\[12 pt]
  {\Large \textbf{Extra practice \ --- \ Randomly walking inventory}} \\[24 pt]
\end{center}

Suppose a company has developed a revolutionary zero-emission airplane, which is in high demand.
Let $I(t)$ be the inventory of planes available to sell at time $t$, with negative values representing a backlog of planes already ordered that still need to be produced.
Ten pre-orders produce a backlog of $I_0 \equiv I(0) = -10$ when production starts at time $t = 0$.
Once production starts, a single plane is produced every $\De t = 3$~weeks.
Every single week there is a 50\% chance of receiving a new order for a plane.

Treat the inventory as a random walk, taking one step every $\De t = 3$~weeks. \\[-30 pt]

\begin{enumerate}[label={(\alph*)}]
  \item What are the possible step sizes $\De I$ and the corresponding probabilities for the random walk described above? % TODO: May want to make sure students get started reliably...
        Compute the resulting mean and variance of the single-step process.

  \item What is the minimum time $t_{\text{min}}$ in which the initial backlog could be cleared, to give $I(t_{\text{min}}) \geq 0$?
        What is the probability the backlog will be cleared in this time?

  \item Using the central limit theorem, what is the expected size of the inventory when $t = 78$~weeks (approximately 18 months)?

  \item Using the central limit theorem and the integral
        \begin{equation*}
          \frac{1}{\sqrt{\pi}} \int_{-\infty}^U e^{-u^2} \d{u} \approx 0.9984  \qquad \mbox{for} \qquad U = 2.082,
        \end{equation*}
        what is the probability the backlog will be reduced, so that $I(t) > I_0$, when $t = 78$~weeks?
\end{enumerate}


\end{document}
% ------------------------------------------------------------------
